Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-10-11 paper/paper.tex

This article gathers three novel contributions.
First, our model encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production \cite{lostanlen2018extended}.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
Secondly, we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.
A comprehensive ablation study of our system demonstrates that all design choices presented hereafter benefit the retrieval precision of our system.
Thirdly, we supplement our quantitative benchmark with some visualizations of time--frequency scattering coefficients in the rate--scale domain, for various typical samples of realistic instrumental playing techniques.
These visualizations confirm the theoretical predictions of previous research \cite{anden2015mlsp}, which relied on simple synthetic sounds such as exponential chirps.
They are also in line with visualizations of the modulation power spectrum in auditory neurophysiology, while offering an accelerated algorithm for scalable feature extraction.

%% TODO: list Sections

%% TODO: mention companion website and open-source software

\section*{Related work}
Timbre involves multiple time scales in conjunction, from a few microseconds for an attack transient to several seconds for a sustained tone.
Therefore, computational models of timbre perception must summarize acoustic information over a long analysis window, typically amounting to $10^{4}$ digital audio samples or more \cite{joder2009taslp}.
Mapping this input to a feature space in which distances denote timbral dissimilarities requires a data-driven stage of dimensionality reduction.
In this respect, the scientific literature exhibits a methodological divide as regards the collection of human-annotated data \cite{siedenburg2016jnmr}.

@@

\section*{Data collection of timbre disimilarity judgments}
\label{sec:survey}

>>>>

This article gathers three novel contributions.
First, our model encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production \cite{lostanlen2018extended}.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
Secondly, we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.
A comprehensive ablation study of our system demonstrates that all design choices presented hereafter benefit the retrieval precision of our system.
Thirdly, we supplement our quantitative benchmark with some visualizations of time--frequency scattering coefficients in the rate--scale domain, for various typical samples of realistic instrumental playing techniques.
These visualizations confirm the theoretical predictions of previous research \cite{anden2015mlsp}, which relied on simple synthetic sounds such as exponential chirps.
They are also in line with visualizations of the modulation power spectrum in auditory neurophysiology \cite{patil2012ploscompbiol}, while offering an accelerated algorithm for scalable feature extraction.

The remaining of this paper is organized as follows. Section \ref{sec:soa} presents an in-depth review of the state of the art in the computational modeling of musical timbre. The perceptual data that allows us to train and evaluate the proposed model is briefly presented in Section \ref{sec:survey}. Extended presentation of the gathering protocol used for collecting this data is presented in Annex \ref{sec:subjective}. The computational model and its core components are described in Section \ref{sec:method}. Their benefits are evaluated in Section \ref{sec:results} and discussed in Section \ref{sec:discussion}\footnote{Reproducible research statement: the implementation of the proposed model and the experimental protocol used to produce the results discussed in this paper are available alongside the anonymized version of the perceptual data: \url{https://github.com/mathieulagrange/paperPlayingTechniquesSimilarityLearning}}.

\section*{Related work}
\label{sec:soa}

Timbre involves multiple time scales in conjunction, from a few microseconds for an attack transient to several seconds for a sustained tone.
Therefore, computational models of timbre perception must summarize acoustic information over a long analysis window, typically amounting to $10^{4}$ digital audio samples or more \cite{joder2009taslp}.
Mapping this input to a feature space in which distances denote timbral dissimilarities requires a data-driven stage of dimensionality reduction.
In this respect, the scientific literature exhibits a methodological divide as regards the collection of human-annotated data \cite{siedenburg2016jnmr}.

@@

\section*{Data collection of timbre similarity judgments}
\label{sec:survey}
